---
title: "Getting started with ggvolc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ggvolc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  dpi = 110,
  message = FALSE,
  warning = FALSE
)
```

`ggvolc` turns the output of a differential-expression (DE) analysis into a
polished, publication-ready volcano plot. It accepts the output of **DESeq2**,
**edgeR**, and **limma** directly, highlights and labels genes of interest,
combines the plot with a `gt` table, and can render an interactive version with
`ggiraph`.

```{r}
library(ggvolc)
```

## The example data

The package ships with two data frames. `all_genes` is a complete DESeq2-style
result table; `attention_genes` is a small subset of genes you might want to
highlight.

```{r}
data(all_genes)
data(attention_genes)

head(all_genes)
```

## A basic volcano plot

Passing a single data frame colours every gene by significance. By default a
gene is called significant when its **adjusted** p-value (FDR) is below `0.05`
and its `|log2FoldChange|` exceeds `1`.

```{r}
ggvolc(all_genes)
```

## Highlighting genes of interest

Supply a second data frame and those genes are drawn with a black outline and
labelled.

```{r}
ggvolc(all_genes, attention_genes, add_seg = TRUE)
```

The `add_seg = TRUE` argument adds dashed guides at the fold-change and
significance thresholds.

## Labelling the top genes automatically

You usually do not want to build a separate data frame just to label a handful
of hits. `label_top` picks the most significant genes for you, and `label_dir`
controls the direction they are drawn from.

```{r}
# the 10 most significant genes overall
ggvolc(all_genes, label_top = 10, add_seg = TRUE, title = "Top 10 hits")
```

```{r}
# the top 8 up- and top 8 down-regulated genes
ggvolc(all_genes, label_top = 8, label_dir = "each",
       add_seg = TRUE, title = "Top 8 up + 8 down")
```

`label_dir` accepts `"both"` (the default), `"up"`, `"down"`, and `"each"`.

## Calling significance on the FDR (or the raw p-value)

`ggvolc()` calls significance on the adjusted p-value (`padj`) by default, which
is the right cutoff for most DE workflows. The y-axis and the significance line
follow the choice, so the plot stays internally consistent. Switch to the raw
p-value with `sig_col = "pvalue"`.

```{r}
ggvolc(all_genes, sig_col = "pvalue", add_seg = TRUE,
       title = "Significance on the raw p-value")
```

`ggvolc()` is also robust to p-values of exactly `0` (which DESeq2 and edgeR can
emit for the strongest genes): rather than dropping those points, their
`-log10` value is capped so they stay near the top of the plot.

## Scaling point size

Point size can encode either the fold change or the significance.

```{r}
ggvolc(all_genes, attention_genes, size_var = "log2FoldChange", add_seg = TRUE)
```

## Adding a gene table

`genes_table()` composes the volcano plot with a `gt` table of the highlighted
genes using `patchwork`. The result is a single object you can save with
`ggplot2::ggsave()`.

```{r, fig.height = 7}
p <- ggvolc(all_genes, attention_genes, add_seg = TRUE)
genes_table(p, attention_genes)
```

## Works with DESeq2, edgeR, and limma

Column names from all three pipelines are detected and mapped automatically, so
you can pass their output straight in. Gene identifiers held in the row names
(as edgeR and limma often do) are promoted to a `genes` column for you.

The package ships `edger_genes`, an example `topTags()`-style table with the gene
identifiers in the row names, so you can try this right away:

```{r}
data(edger_genes)
head(edger_genes, 3)

ggvolc(edger_genes, label_top = 8, add_seg = TRUE, title = "edgeR input")
```

| Pipeline | Fold change | p-value | adjusted p | expression |
|----------|-------------|---------|------------|------------|
| DESeq2 (`results()`) | `log2FoldChange` | `pvalue` | `padj` | `baseMean` |
| edgeR (`topTags()`) | `logFC` | `PValue` | `FDR` | `logCPM` |
| limma (`topTable()`) | `logFC` | `P.Value` | `adj.P.Val` | `AveExpr` |

## An interactive volcano

With the optional [`ggiraph`](https://davidgohel.github.io/ggiraph/) package
installed, `interactive = TRUE` returns a widget where hovering a point reveals
the gene name and its statistics.

```{r, eval = requireNamespace("ggiraph", quietly = TRUE)}
ggvolc(all_genes, attention_genes, interactive = TRUE)
```

```{r, eval = !requireNamespace("ggiraph", quietly = TRUE), echo = FALSE, results = "asis"}
cat("_Install `ggiraph` to render the interactive version:_ `install.packages(\"ggiraph\")`.")
```
